Innovation and elaboration on the avian tree of life - supplementary materials

### 2022-10-20 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''. At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}. The presence of distinct morphological forms suggests instead that species or clades may evolve away from the line of least resistance. At a macroevolutionary level, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal, renaud2006conserved}. The extent to which elaboration and innovation contribute to phenotypic diversity among species and clades is critical to our understanding of the origins and maintenance of biodiversity across scales from the micro- to macro and mega-evolutionary. Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks among species and clades. Our analyses show that elaboration is a common feature at all scales, consistent with theory.
We also find that innovation is a common and major contributor to avian morphological diversity among clades.
Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

@@

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}. Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}. Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales. Recent evidence even suggests that this stability can extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}. However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic divergence is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}. The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the directions of phenotypic evolution across scales where macroevolutionary divergence along lines of least resistance gives way to mega-evolutionary divergence leading to distinct phenotypes or directions of evolution. Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

% paragraph 2:  introducing the Endler view.
 Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two  routes to diversity that we suggest can help to illuminate this challenge. Specifically, Endler et al. refer to evolution of new phenotypes along the same axis as \textit{elaboration}. This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution. In contrast, they refer to evolution away from the major axis as \textit{innovation}. More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.  For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022} are consistent with specific innovation. Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales. These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of: i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of divergence (specific innovation), and ii) evolution along versus away from major axes of divergence. The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
 Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.

% paragraph 3: how we measure this (glmm) on the birds data + nestedness
\subsection{Modelling nested covariances}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
1) large, multivariate datasets to allow exploration of trait covariances at different scales,
2) reliable and efficient computational methods to estimate the major axes of phenotypic divergence,
and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale. To meet the first challenges we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}). To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}). In order to provide a baseline on which to calculate elaboration and innovation we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}). Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components. From these models we can calculate major axes of divergence as the leading eigenvector in trait space. We regard the major axis of the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution. The major axis can be defined at any scale (i.e. for all species in the phylogeny, or for all species within any consistent clade). In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of divergence) for i) the phylogenetic partition of all species, ii) all species within each of nine clades mapping approximately to superorders, and iii) all species within each of XX clades mapping to taxonomic orders. These nested partitions of multivariate trait space provide the basis for all subsequent analyses.

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We then measure innovation and elaboration as the multilinear algebra projection of each species onto their nested major axis of variation.
Here are the results for each species. \ref{Fig:phylogeny}
There seems to be a nested structure of the variation in novelty %TODO: needs to introduce nestedness; %TODO: needs to introduce novelty.
There is more variation the deeper you baseline comparison in the tree is, suggesting that novelty is relative to the baseline comparison.
Also, the majority of that novelty is through elaboration.
We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
However, we also found various combinations of elaboration and innovation across birds.
For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
\textbf{But there is a paradox here:}
If most organisms evolve incrementally along major axis of variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

\begin{figure}[!htbp]
\centering
    \includegraphics[width=0.9\textwidth]{Figures/median_novelty_distributions.pdf}
\caption{THIS FIGURE SHOULD BE EMBEDED IN THE ANUS FIGURE: showing the novelty grouped into two different ways}
\label{Fig:phylogeny_2}
\end{figure}

@@

\subsection{Multiple routes to specific innovation throughout avian evolutionary history}
% Paragraph 5: paradox + methods for comparing ellipses
Although the adaptive radiation theory implies that phenotypic divergence arises primarily along the major axis of variation \cite{marroig2005size}, there has been many contrasting observations in the literature highlighting major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the phenotypic major axis within different clades compared to the global phylogenetic major axis.
We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis (Fig. \ref{Fig:ellipses} - Fig. \ref{Fig:orthogonality}).
Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the global phylogenetic major axis and each super order (Fig. \ref{Fig:ellipses}a) and order (Fig. \ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig. \ref{Fig:orthogonality}).
The median angle of the major axis for subclades approaches orthogonality, differing from the phylogenetic major axis by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
Comparisons of orientation among subclades (i.e. orders within superorders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.

% Paragraph 5+: Some elluipses stuff with bird names.
However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
Clades (or species) can be both elaborators and innovators.
To measure innovation and elaboration we translated the major axes of each clade's (i.e. superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g. Cuculiformes; Fig. \ref{Fig:ellipses}), low elaboration and high innovation (e.g. Galliformes; Fig. \ref{Fig:ellipses}), high elaboration and low innovation (e.g. Coraciiformes; Fig. \ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g. Podicipediformes; Fig. \ref{Fig:ellipses}).
These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
This suggests that divergence away from line of least evolutionary resistance (i.e. the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.

This apparent lack of deep-time evolutionary constraint in bird beaks is supported by variation in both the dominant direction of evolution within clades and their relative dimensionality.
Although the major axis of the global phylogenetic component aligns closely with the first dimension (70.96\% of the phylogenetic major axis is aligned on PC1; Fig. \ref{Fig:ellipses}) of the raw shapespace, no superorder, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), mainly align with PC1 in shapespace Fig. \ref{Fig:ellipses}-bar plots).
These patterns are similar (albeit more contrasted) when looking at sub-orders and families within passeriformes half off the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with PC1 (Fig. S\ref{fig_ellipses_passeriformes}).
Moreover, we found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the global phylogenetic partition.
For example, Accipitriformes (hawks and allies) have major axes that mainly align with the second dimension of the shapespace suggesting uniformity in their beak shape components, whereas the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.

@@

\noindent \textbf{Figure \ref{Fig:ellipses}:}
Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each superorder (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
We scaled the ellipses so the length of the major axis of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
The first inset ellipse plot shows the positions of all superorder ellipses relative to the all-bird phylogenetic ellipse.
Subsequent inset plots show the results for each superorder. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.

@@

\noindent \textbf{Figure \ref{Fig:orthogonality}:}
Orthogonality of each clade's major axis compared to their parent or parent parent's clade.
Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
Dots represent the clades median orthogonality and the lines their 95\% CI.
No super order's major axis is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
Only two out of six super-orders; 6/25 orders relative to their super order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.

% Paragraph 7: putting these results together
\subsection{Species elaborate, clades innovate}

>>>>

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''. At macro- and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}. The presence of distinct morphological forms suggests instead that species or clades may evolve away from the line of least resistance. At a macroevolutionary level, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal, renaud2006conserved}. The extent to which elaboration and innovation contribute to phenotypic diversity among species and clades is critical to our understanding of the origins and maintenance of biodiversity across scales from the micro- to macro and mega-evolutionary. Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks among species and clades. Our analyses show that elaboration is a common feature at all scales, consistent with theory.
We also find that innovation is a common and major contributor to avian morphological diversity among clades.
Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

@@

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

% paragraph 2:  introducing the Endler view.
Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
Specifically, Endler et al. refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
In contrast, they refer to evolution away from the major axis of phenotypic variation as \textit{innovation}.
More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022} are consistent with specific innovation.
Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
These core concepts provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of phenotypic variation (specific innovation),
and ii) evolution along versus away from major axes of phenotypic variation.
The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.

% paragraph 3: how we measure this (glmm) on the birds data + nestedness
\subsection{Modelling nested trait covariances}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
1) large, multivariate datasets to allow exploration of trait covariances at different scales,
2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
To provide a baseline from which to calculate elaboration and innovation (i.e. the major axis of phenotypic variation) we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components. 
From these models we can calculate major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix in trait space.
We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any monophyletic clade).
To assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of phenotypic variation) for
i) the entire phylogeny of all species included in the data (hereafter the "whole tree" level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent analyses.

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species specific measure of random innovation and elaboration. To do this we estimated the multilinear algebraic projection of each species phenotype onto the three (whole tree, super-order, and order) major axis of phenotypic variation. Phylogenetic distributions of random innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or whole tree level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
However, the extent of both elaboration and random innovation tends to reduce as we move from whole tree to within order comparison.
Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
The whole tree's major axis of phenotypic variation primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.

Overall we found that, as expected, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}. We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
However, we observe various combinations of elaboration and random innovation across birds. Examples of high innovation and/or elaboration are often clustered within clades. For example, compared to the whole tree major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation. In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts) in the order Apodiformes and the distantly related but convergent Hirundidae (swallows and martins) have exceptionally wide (relative to length) beaks and display high levels of both elaboration and specific innovation.
The dominance of elaboration when measured at the species level is consistent with idea of evolution along lines of least evolutionary resistance. However, the frequent clustering of species that elaborate, innovate, or both, again highlights the likelihood that major axes of phenotypic variation are not stable among clades.
This leads us to ask: if most organisms evolve incrementally along major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

@@

\subsection{Multiple routes to specific innovation throughout avian evolutionary history}

To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and specific innovation of different clades.
To do this we translated the major axes of phenotypic variation of each clade's (i.e. super-orders and orders) phylogenetic variance-covariance matrices onto the whole tree's major axis of phenotypic variation so that they shared the same origin in the shapespace.
We then used multilinear algebra to project a focal clade's major axis of phenotypic variation onto the whole tree's major axis of phenotypic variation  (Fig.S\ref{Fig:cheat_sheet}).
We interpret the measured projection (distance onto the phylogenetic major axis of phenotypic variation) as the clade's elaboration score and the measured rejection (distance from the major axis of phenotypic variation) as the clade's innovation score.
These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
We found that for most clades the phenotypic major axis of phenotypic variation is not aligned with the whole tree's major axis of phenotypic variation , (Fig. \ref{Fig:ellipses}).
Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, specific innovation is a more common generator of morphological diversity than elaboration in deep time.
However, there is substantial variation in elaboration and innovation across clades including relatively low elaboration and innovation (e.g. Cuculiformes; Fig. \ref{Fig:ellipses}), low elaboration and high innovation (e.g. Galliformes; Fig. \ref{Fig:ellipses}), high elaboration and low innovation (e.g. Coraciiformes; Fig. \ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g. Podicipediformes; Fig. \ref{Fig:ellipses}).
These patterns hold across scales including within super-orders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
This suggests that there is no deep time analogue of the evolutionary line of least evolutionary resistance and that across scales there is a remarkable flexibility in the routes to innovation.

@@

\noindent \textbf{Figure \ref{Fig:ellipses}:}
Panel A) shows  ellipses representing the scaled median posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the all-birds phylogenetic component of the models (grey ellipses)
We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipses is the same length as that of the global ellipse (in eight dimensions).
The first inset ellipse plot shows the positions of all super-order ellipses relative to the all-bird phylogenetic ellipse.
Subsequent inset plots show the results for each super-order. Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
Note that these scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation scores for each plot. Panel B) shows  the results for each order.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.

This variability in the routes to innovation implies a distinct lack of deep-time evolutionary constraint in the dominant axes of evolution within clades.
Although the whole tree's major axis of phenotypic variation aligns closely with the first dimension (70.96\% of the whole tree's major axis of phenotypic variation is aligned on PC1; Fig. \ref{Fig:ellipses}) of the raw shapespace, no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), follow this trend Fig. \ref{Fig:ellipses}-bar plots). This also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes PC1 (Fig. S\ref{fig_ellipses_passeriformes}).
In addition to the variable alignment of clade major axes of phenotypic variation, we further found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis) or higher-dimensional than expected from the whole tree variance-covariance matrix.
For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the whole tree but uniformity in their beak shape components within the clade. In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
These observations illustrate multiple routes to innovation and imply that such innovation could arise at any time throughout avian history.
However, although we see repeated innovation at the clade level, providing a new hypothesis to explain the expansion of avian morphological space, these analyses do not allow us to distinguish whether reorientation arises through abrupt changes or more gradual realignment of species in trait space.

\subsection{Species elaborate, clades innovate}

% Paragraph 5: paradox + methods for comparing ellipses
Although adaptive radiation theory implies that phenotypic divergence arises primarily along the major axis of phenotypic variation \cite{marroig2005size}, there have been many contrasting observations in the literature highlighting major jumps in phenotypic space at the megaevolutionary level \cite{puttick2014high,cooney2017mega,venditti2011multiple}.
We used posterior distributions of variance-covariance matrices derived from our pGLMMs to calculate differences in the major axis of phenotypic variation within different clades compared to the whole tree's of phenotypic variation.
We find overwhelming evidence that the orientation of evolution within clades is markedly different from - and frequently orthogonal to - the major phylogenetic axis of phenotypic variation (Fig. \ref{Fig:ellipses} - Fig. \ref{Fig:orthogonality}).
Despite the limitations representing eight-dimensional space in 2D, our plots of the median elliptical representation of the phenotypic variance-covariance matrix for the whole tree's major axis of phenotypic variation and each super order (Fig. \ref{Fig:ellipses}a) and order (Fig. \ref{Fig:ellipses}b) highlight the variation in the orientation of phenotypic evolution among clades (Fig. \ref{Fig:orthogonality}).
The median angle of the major axis of phenotypic variation for subclades approaches orthogonality, differing from the phylogenetic major axis of phenotypic variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$).
Comparisons of orientation among subclades (i.e. orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary level and are no more likely to occur along any one axis than another.
This differs from previous inference on a subset of our data that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
These results suggest that the line of least evolutionary resistance is itself evolutionarily labile and are consistent with the idea that morphological divergence may be less constrained in deep-time than is often assumed \cite{venditti2011multiple}.

@@

\noindent \textbf{Figure \ref{Fig:orthogonality}:}
Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent parent's clade.
Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines represent respectively an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
Dots represent the clades median orthogonality and the lines their 95\% CI.
No super-order's major axis of phenotypic variation is parallel to the whole phylogeny (red lines and circles) and 12/25 orders are at least half orthogonal to their super-order (green lines and circles).
These results are even clearer when comparing them to the whole phylogeny (green dashed lines and open circles) where only 3/25 are not at least half orthogonal.
Furtheremore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
For each clade we also measured the posterior propbabilityof each clade's orientation being different from their parent's clade or the whole phylogeny relative to their sample size and sd.
The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
Only two out of six super-orders; 6/25 orders relative to their super-order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
This is the result of a variation of sample size (n) and standard variation (sd) between clades. 
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.

% Paragraph 7: putting these results together
